P3692R0: How to Avoid OOTA Without Really Trying

### 2025-05-22 paper/ootavacuous.tex
ootavacuous; Alan Stern May 19 2025 updates
Message ID: 87f55604-c3b6-4f4e-960d-e83026c5ea18@rowland.harvard.edu

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly recommends against them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no ``out-of-thin-air'' values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.
(The as-if rule permits those optimizations where the optimized code's
observable behavior is the same as if it were unoptimized.)

@@

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
T0 and T1.
(Roughly speaking, there is a semantic dependency from a given load to
a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored, change the target address of
the store, or prevent the store from occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	A similar restriction applies to compiler-based speculation.}
the stores on line~2 cannot take place until T0's and T1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
provided a decision procedure for distinguishing between reordering and
OOTA, using a perturbation method based on the insight that OOTA cycles
are fixed-point computations
In other words, an OOTA cycle must produce the same value that it
consumes.

@@

Other researchers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA,MinkiCho2021PromisingSemantics,SohamChakraborty2019OOTAeventStructure,JeehoonKang2017PromisingSemantics,JeanPichonPharabod2016OOTA},
but none of these have been accepted by C++ implementers, in part due
to these models requiring consideration of multiple executions.
As of 2024, Mark Batty is working on modular relaxed dependencies,
which may have the added benefit of checking compiler
optimizations~\cite{JayRichards2024SymbolicMrder}, which is a portion
of the OOTA problem that is out of scope for this paper.\footnote{
	But please see Section~\ref{sec:Merging Quasi-Volatile Loads}.}

Moiseenko et al.~avoid cycles in po $\cup$
rf~\cite{EvgeniiMoiseenko2025reexecution}, which might permit
more optimizations than forcing prior relaxed loads to be ordered
before subsequent relaxed stores, but which the authors characterize
as ``(almost) a per-execution model''.
It remains to be seen whether this is close enough to a per-execution
model to be accepted by C++ implmenters.
\emph{@@@ I am not fully confident of my characterization of this paper.}

@@

Our approach applies real-world hardware constraints (which are further
constrained by long-standing laws of physics) to the problem of avoiding
OOTA cycles.
We do this by augmenting the C++ abstract machine with these constraints,
which of course limits the applicability of our results to C++ programs
compiled and run on real-world computer systems.
Given that a quarter century of OOTA research based on pure unaugmented
C++ abstractions has resulted in (at best) incremental progress on the
one hand, and that real-world computer systems comprise an exceedingly
important part of the overall OOTA problem on the other, our results
are valuable despite these limitations.
In the wise words of Peter J.~Denning, ``Maybe we need to occasionally
descend from the high clouds of our abstractions to the concrete earthy
concerns of everyday life''~\cite{PeterJDenning2025Abstraction}.

The key observations that enable us to apply these constraints are:
(1)~Communicating data within a computer system takes time,
(2)~Executing instructions that compute an output value that semantically
depends on given input values also takes time,
and
(3)~Each link in an OOTA cycle consists of either data communication
or computation of an output that semantically depends on an input.
Because closing an OOTA cycle comprising a sequence of commmunications
and computations requires this sequence to end at the same time as it
starts, OOTA cycles cannot form in those C++ implementations that are
within the scope of this paper.
A great advantage of our approach is that we are \emph{not} required to
precisely define the boundaries of sdep.
After all, although compilers are forbidden from breaking semantic
dependencies, they are by no means required to break non-semantic
dependencies.

These simple-seeming observations conceal considerable complexity,
and so much of the remainder of this paper deals with corner cases that
arise when carefully considering semantic dependencies, executions, C++
compilers, and hardware properties.

In short, this paper's contribution is to show how these
constraints prevent real-world C++ implementations doing per-thread
analysis from exhibiting OOTA behavior, and not with minimal
cost~\cite{PiezhaoOu2018OOTApapl}, but rather with no cost at
all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
\emph{@@@ Are these the right qualifiers?}

\section{OOTA and Semantic Dependencies (sdep)}
\label{sec:OOTA and Semantic Dependencies (sdep)}

@@

Semantic dependencies form only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to argue that the store must precede
the load in global time and combine this with the intuitive notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 (``Out-of-Thin-Air Execution is Vacuous'').
where sdep is the set of semantic dependencies within each thread and
rf is the set of store-to-load ``reads-from'' links, whether internally
within a thread (rfi) or externally between threads (rfe).\footnote{
	Additional background on temporal distinctions between coe, fre,
	and rfe maybe found elsewhere~\cite[Appendix A]{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.}

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the intuitive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later in global time than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction later on.

@@

Here we consider some of the complexities inherent to semantic
dependencies.
It is important to distinguish semantic dependencies from syntactic
dependencies.
Although every semantic dependency is also a syntactic dependency, many
syntactic dependencies can be optimized (or ``broken''), as will be seen
in the following sections.
Such syntactic dependencies are not semantic dependencies.

@@

Some discussions of semantic dependencies assume that they
are strictly functions of the source code.
This assumption is not always valid, in fact, we will see that many
instances of semantic dependency must be considered functions of
particular executions.
Consider for example:
\begin{quote}
\begin{verbatim}
z = x * y;
\end{verbatim}
\end{quote}

@@

Since we can no longer regard sdep as always relating a single load to a store,
the notion of a cycle involving sdep appears problematic.
We are forced to change our definition of an OOTA cycle again;
we will say that an execution is an instance of OOTA if in that execution:
\begin{quote}
	There are stores $W_0,$ \ldots,~$W_m$,
	where each $W_i$ semantically depends on a set of loads
	$\{R_{i,0},$ \ldots,~$R_{i,n_i}\}$,
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}
This can make OOTA more complicated than a simple cycle but
we will continue to refer to ``OOTA cycles'' out of habit.
Note that this new definition includes and generalizes the earlier
``cycle in sdep $\cup$ rfe'' definition.

@@

We have seen several examples showing that semantic dependencies can
be more complicated than might first appear, varying according to the
execution and even the implementation.
This raises several questions, of which the first is:
What exactly is an execution?

@@

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior)
This requirement is the standard's ``as-if'' rule.
This means:
(1)~The program's output must be the same as that of the abstract execution;
(2)~Volatile accesses ``are evaluated strictly according to the
rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
(3)~[A condition on the timing and interleaving of input and output,
which does not matter for our purposes].
We will say that the hardware execution realizes the abstract execution,
or, equivalently, that the abstract execution is \emph{realized by}
the hardware execution.

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code:
% Constructed by hand.
\begin{quote}
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}
\end{quote}
The loads previously on line~1 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs, even if relaxed loads are ordered
before relaxed stores!

@@

If all the stores in the OOTA cycle are omitted then all the reads must
also be omitted, meaning that the entire cycle has been optimized away.
We conjecture that in this situation there must be another abstract
execution which has the same observable effects as $A$ and is also
realized by $H$, but in which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$H$, whether or not there was an OOTA cycle.
We therefore declare OOTA cycles in which all accesses are omitted to
be \emph{trivial}.

@@

We informally define a \emph{witness} to be another execution in which
at least one of the loads obtains a different value but all else is the
same as far as possible, and yet the store acts differently.

@@

Here we consider some general questions related to our definition of
semantic dependency.
Additional information is available in a companion technical report,
which is not cited in order to maintain anonymity.
% P3064R2 (``How to Avoid OOTA Without Really Trying'')
% https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2024/p3064r2.pdf

@@

Although this is unexpected, and while we don't recommend this sort of
code transformation, we cannot say that this conclusion is definitely
wrong, because one's intuitive notions of semantic dependency are not
always clear in cases involving multiple loads from the same variable.
% Can we instead say that transforming the code might require that the
% witness also be transformed?
In the unlikely event that this transformation proves useful, there
will be a community process that determines whether the value of this
transformation is worth the added conceptual load on developers.
Until that time, pure mathematics cannot help us because it is simply
impossible to say whether or not there is a semantic dependency from
\co{y} to \co{z} without reference to a specific implementation.

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and just-in-time
compilers (JITs) on
one hand, and to link-time optimization (LTO) on the other;
(2)~Relax quasi-volatile semantics to permit limited load speculation,
for example, use of conditional-move instructions in cases where the
source code conditionally consumes the results of the atomic load;
(3)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(4)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(5)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(6)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(7)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.

@@

Sensible though this recommendation might be, it can unnecessarily raise
anxiety levels of C++ implementers.
After all, what exactly do they need to do in order to follow that
recommendation?
One way to reduce their anxiety levels would be to add something like
the following sentence to this recommendation:
\begin{quote}
	Compiler-based implementations whose binaries run on conventional
	hardware are guaranteed not to compute out-of-thin-air values as
	long as they restrict themselves to thread-at-a-time analysis,
	valid optimizations, and user programs that are free of undefined
	behavior.
\end{quote}

@@

The conclusion to be drawn from this work is that in this context,
avoiding OOTA cycles requires no changes to the standard,
to current implementions, or to user practices for portable code.
% This work also demonstrates the need for caution when optimizing
% atomic operations.

>>>>

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly recommends against them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no ``out-of-thin-air'' values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.
(The as-if rule permits optimizations where the optimized code's
observable behavior is the same as if it were unoptimized.)

@@

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
T0 and T1.
Roughly speaking, there is a semantic dependency from a given load to
a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored, change the target address of
the store, or prevent the store from occurring at all.}
(A \emph{syntactic} dependency, by contrast, exists whenever the value
of the load appears in the source code as part of the computation
leading up to the store, whether it affects the store or not.
Only semantic dependencies are relevant to OOTA.)
Here the semantic dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.

Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	A similar restriction applies to compiler-based speculation.}
the stores on line~2 cannot take place until T0's and T1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
provided a decision procedure for distinguishing between reordering and
OOTA, using a perturbation method based on the insight that OOTA cycles
are fixed-point computations
(in other words, an OOTA cycle must produce the same value that it consumes).

@@

Other researchers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA,MinkiCho2021PromisingSemantics,SohamChakraborty2019OOTAeventStructure,JeehoonKang2017PromisingSemantics,JeanPichonPharabod2016OOTA},
but none of these have been accepted by C++ implementers, in part due
to these models requiring consideration of multiple executions.
As of 2024, Mark Batty is working on modular relaxed dependencies,
which may have the added benefit of checking compiler
optimizations~\cite{JayRichards2024SymbolicMrder}, a portion
of the OOTA problem that is out of scope for this paper.\footnote{
	But please see Section~\ref{sec:Merging Quasi-Volatile Loads}.}

Moiseenko et al.~avoid cycles in po $\cup$
rf~\cite{EvgeniiMoiseenko2025reexecution}, which might permit
more optimizations than forcing prior relaxed loads to be ordered
before subsequent relaxed stores, but which the authors characterize
as ``(almost) a per-execution model''.
It remains to be seen whether this is close enough to a per-execution
model to be accepted by C++ implementers.
\emph{@@@ I am not fully confident of my characterization of this paper.}

@@

Our approach applies real-world hardware constraints (including some
imposed by long-standing laws of physics) to the problem of avoiding
OOTA cycles.
We do this by relating the C++ abstract machine to hardware executions,
which of course limits the applicability of our results to C++ programs
compiled and run on real-world computer systems.
Given that a quarter century of OOTA research based on pure unaugmented
C++ abstractions has resulted in (at best) incremental progress on the
one hand, and that
compiler-based real-world computer systems comprise an exceedingly
important part of the overall OOTA problem on the other, our results
are valuable despite these limitations.
In the wise words of Peter J.~Denning, ``Maybe we need to occasionally
descend from the high clouds of our abstractions to the concrete earthy
concerns of everyday life''~\cite{PeterJDenning2025Abstraction}.

The key constraints mentioned above are:
(1)~Communicating data within a computer system takes time, and
(2)~Executing instructions that compute an output value that semantically
depends on given input values also takes time.
Because each link in an OOTA cycle consists of either data communication
or computation of an output that semantically depends on an input,
and because closing an OOTA cycle comprising a sequence of commmunications
and computations requires this sequence to end at the same time as it
starts, OOTA cycles cannot form in those C++ implementations that are
within the scope of this paper.
A great advantage of our approach is that we do \emph{not} need to define
semantic dependency purely at the level of C++ source code;
rather, we are able to utilize information already provided by existing
hardware execution models.

These simple-seeming ideas conceal considerable complexity,
and much of the remainder of this paper deals with corner cases that
arise when carefully considering semantic dependencies, executions, C++
compilers, and hardware properties.

In short, this paper's contribution is to show how these
constraints allow real-world C++ implementations doing per-thread
analysis to avoid OOTA behavior easily,
%and not with minimal cost~\cite{PiezhaoOu2018OOTApapl},
in many cases with no cost at
all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
\emph{@@@ Are these the right qualifiers?}

\section{OOTA, Semantic Dependencies, and Reads-From}
\label{sec:OOTA, Semantic Dependencies, and Reads-From}

@@

Semantic dependencies form only one type of link in an OOTA cycle;
the other type extends from a given store to a load that returns the
value stored.
In principle, the time required by each of these links should prevent
OOTA cycles from occurring.
This idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 (``Out-of-Thin-Air Execution is Vacuous'').
where sdep is the set of semantic dependencies within each thread and
rf is the set of store-to-load ``reads-from'' links, whether internally
within a thread (rfi) or externally between threads (rfe).\footnote{
	Additional background on temporal distinctions between coe, fre,
	and rfe maybe found elsewhere~\cite[Appendix A]{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.}

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the naive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction later on.

@@

Here we consider some of the complexities inherent to semantic
dependencies.

@@

Some discussions of semantic dependencies assume that they
are strictly properties of the source code.
This assumption is not always valid; in fact, many
instances of semantic dependency vary according to the details of
particular executions.
Consider for example:
\begin{quote}
\begin{verbatim}
z = x * y;
\end{verbatim}
\end{quote}

@@

Since we can no longer regard sdep as always relating a single load to a store,
the notion of a cycle involving sdep appears problematic.
We are forced to change our definition of OOTA again;
we will say that an execution is an instance of OOTA if in that execution:
\begin{quote}
	There are stores $W_0,$ \ldots,~$W_m$,
	where each $W_i$ semantically depends on a set of loads
	$\{R_{i,0},$ \ldots,~$R_{i,n_i}\}$,
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}
This can make OOTA more complicated than a simple cycle but
we will continue to refer to ``OOTA cycles'' out of habit.
Note that this new definition includes and generalizes the earlier
``cycle in sdep $\cup$ rfe'' definition.

@@

\medskip

We have seen several examples showing that semantic dependencies can
be more complicated than might first appear, varying according to the
execution and even the implementation.
This raises several questions, of which the first is\dots

@@

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior).
This requirement is the standard's ``as-if'' rule.
It means:
(1)~The program's output must be the same as that of the abstract execution;
(2)~Volatile accesses ``are evaluated strictly according to the
rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
(3)~[A condition on the timing and interleaving of input and output,
which does not matter for our purposes].
We will say that the hardware execution \emph{realizes} the abstract execution,
or, equivalently, that the abstract execution is \emph{realized by}
the hardware execution.

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code:
% Constructed by hand.
\begin{quote}
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}
\end{quote}
The loads previously on line~1 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs, even if relaxed loads are ordered
before relaxed stores!
(In fact, the transformed program has no loads at all, relaxed or otherwise.)

@@

If all the stores in the OOTA cycle are omitted then all the reads must
also be omitted, meaning that the entire cycle has been optimized
out of existence.
We conjecture that in this situation there must be another abstract
execution which has the same observable effects as $A$ and is also
realized by $H$, but in which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$H$, whether or not there was an OOTA cycle.
We therefore declare OOTA cycles in which all accesses are omitted to
be \emph{trivial}.

@@

Informally, a \emph{witness} is another execution in which
at least one of the $R_i$ loads obtains a different value but all else is the
same as far as possible, and yet the store $W$ acts differently.

@@

Here we consider some general questions related to our definition of
semantic dependency.
Additional information is available in P3064R2 (``How to Avoid OOTA Without
Really Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.

@@

Although this is unexpected, and while we don't recommend this sort of
code transformation, we cannot say that this conclusion is definitely
wrong, because one's intuitive notions of semantic dependency are not
always clear in cases involving multiple loads from the same variable.
% Can we instead say that transforming the code might require that the
% witness also be transformed?
In the unlikely event that this transformation proves useful, there
will be a community process that determines whether the value of this
transformation is worth the added conceptual load on developers.
Until that time, pure mathematics cannot help us simply because it is
impossible to say whether or not there is a semantic dependency from
\co{y} to \co{z} without reference to a specific implementation.

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and just-in-time
compilers (JITs) on
one hand, and to link-time optimization (LTO) on the other;
(2)~Relax quasi-volatile semantics to permit limited load speculation
(for example, hoisting loads out of the bodies of conditionals);
(3)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(4)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(5)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(6)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(7)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.

@@

Sensible though this recommendation might be, it can unnecessarily raise
anxiety levels of C++ implementers.
After all, what exactly do they need to do in order to follow that
recommendation?
One way to reduce their anxiety levels would be to add something like
the following sentence to this recommendation:
\begin{quote}
	Compiler-based implementations whose binaries run on conventional
	hardware are guaranteed not to compute out-of-thin-air values in
	programs that are free of undefined behavior,
	as long as they restrict themselves to thread-at-a-time analysis
	and do not invent or duplicate atomic accesses.
\end{quote}

@@

The conclusion to be drawn from this work is that in this context,
avoiding OOTA cycles requires no changes to the standard
or to user practices for portable code, and only minimal changes to
implementations.
% This work also demonstrates the need for caution when optimizing
% atomic operations.
